---
title: "Identification of a novel subgroup of endometrial cancer patients with loss of thyroid hormone receptor expression and improved survival"
author: "Daniel G. Piqué, John M. Greally, and Jessica C. Mar"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmdformats::readthedown:
    highlight: kate
    number_sections: true
    includes:
      after_body: footer.html
---



# Introduction

# Setup and download data
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```


## Load in the libraries

```{r}
#library(TCGAbiolinks)  #uncomment if running for first time
library(SummarizedExperiment)
library(RColorBrewer)
library(here)
library(mclust)
library(biomaRt)
library(tidyverse)
library(janitor)
library(survival)
library(ggdendro)
library(limma)
library(fgsea)
library(ggrepel)
library(patchwork)
library(receptLoss)
source(here("scripts", "2019_endomet_smry_suppl.R"))
```

## Download and filter datasets 

**mRNA expression**

* TCGA-UCEC - Endometrial adenocarcinoma
* TCGA-BRCA - Breast adenocarcinoma
* TCGA-LUAD - Lung squamous adenocarcinoma
* TCGA-LUSC - Lung squamous cell carcinoma
* TCGA-OV - Ovarian adenocarcinoma

**microRNA expression**

* TCGA-UCEC - Endometrial adenocarcinoma

We use the R package [`TCGAbiolinks`](https://www.bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html){target="_blank"} to download the expression and clinical data from these cancer types in GDC [data release version 13.0](https://docs.gdc.cancer.gov/Data/Release_Notes/Data_Release_Notes/#data-release-130). 

```{r}

project_types <- c("TCGA-UCEC", "TCGA-BRCA", "TCGA-LUAD", "TCGA-LUSC")

if(!file.exists(here("data", paste0(project_types, "-exp.rds")))){
  ## download the data for each of the 5 cancer types
  ## uncomment the GDCdownload and saveRDS lines if running for first time
  eq <- GDCquery(project = project_types,
                  data.category = c("Transcriptome Profiling"),
                  data.type = c("Gene Expression Quantification"),
                  workflow.type = "HTSeq - FPKM")
  #GDCdownload(eq)
  
  ## prepare and save the datasets
  for(i in 1:length(project_types)){
    eq <- GDCquery(project = project_types[i],
                  data.category = c("Transcriptome Profiling"),
                  data.type = c("Gene Expression Quantification"),
                  workflow.type = "HTSeq - FPKM")
    #GDCdownload(eq)
    e <- GDCprepare(eq, save=FALSE)
    #saveRDS(e, file=here("data",paste0(project_types[i], "-exp.rds")))
    print(project_types[i])
  }
  
  #prepare and save the miRNA datasets
  mi <- GDCquery(project = "TCGA-UCEC",
                 workflow.type = "BCGSC miRNA Profiling",
                 data.category = c("Transcriptome Profiling"),
                 data.type = c("miRNA Expression Quantification"))
  #GDCdownload(mi)
  m <- GDCprepare(mi, save=FALSE)
  #saveRDS(m, file=here("data", "TCGA-UCEC-mirna.rds"))

}

```


## Load and clean the endometrial cancer expression data
```{r}

e <- readRDS(here("data", "TCGA-UCEC-exp.rds")) %>% clean_dataset()
en <- e$en
e <- e$et
  
esr1 = "ENSG00000091831"
ar = "ENSG00000169083"
pr = "ENSG00000082175"

```



## Perform PCA and look for hospital and sequencing plate bias

```{r}
pca <- prcomp(t(assay(e)), center = T, scale. = T)
```


### Add centroids (one for each hospital) overlayed on samples

```{r}
pca.centroids <- aggregate(data.frame(pca$x)[,1:10], list(Type = colData(e)$hosp), mean)

pca.m.hosp <- reshape2::melt(cbind(data.frame(pca$x), hosp=colData(e)$hosp),
                             id.vars=c("PC1", "PC2","hosp"))
pca.centroids.hosp <- aggregate(data.frame(pca$x)[,1:10], list(Type = colData(e)$hosp), mean)
pca.centroids.hosp.m <- reshape2::melt(pca.centroids.hosp, id.vars=c("Type","PC1", "PC2"))

ggplot(pca.centroids.hosp.m, aes(x=PC1, y=PC2, color=Type)) + 
  geom_point(data=pca.m.hosp, aes(PC1, PC2, color=hosp, alpha=0.2), 
             alpha=0.1, size=0.5) +
  centroid_plot_elem()
```

### Add centroids (one for each sequencing plate) overlayed on samples

```{r}
pca.m.plate <- reshape2::melt(cbind(data.frame(pca$x), seqPlate=colData(e)$seqPlate), 
                              id.vars=c("PC1", "PC2","seqPlate"))
pca.centroids.plate <- aggregate(data.frame(pca$x)[,1:10], list(Type = colData(e)$seqPlate), mean)
pca.centroids.plate.m <- reshape2::melt(pca.centroids.plate, id.vars=c("Type","PC1", "PC2"))

ggplot(pca.centroids.plate.m, aes(x=PC1, y=PC2, color=Type)) + 
  geom_point(data=pca.m.plate, aes(PC1, PC2, color=seqPlate, alpha=0.2),
             alpha=0.1, size=0.5) + 
  centroid_plot_elem()
```

We can see that there are no centroids that are visible outliers among either hospitals or sequencing plate. This suggests that the variability observed in the sequencing data is not due to differences in hospitals where the samples were taken from or to differences in the plate that each sample was sequenced on.

## Check and fix BMI clinical data in endometrial cancer
```{r}

colData(e)$bmi <- colData(e)$weight / (colData(e)$height/100)^2
plot(colData(e)$weight, colData(e)$height)

#are there any bmi outliers?
hist(colData(e)$bmi)
bmiOutlIdx <- which(colData(e)$bmi > threeSdAboveMean(mean(colData(e)$bmi, na.rm=T), sd = sd(colData(e)$bmi, na.rm=T)))

#are there any height outliers?
hist(colData(e)$height)
heightOutlIdx <- which(colData(e)$height < threeSdBelowMean(mean(colData(e)$height, na.rm=T), sd = sd(colData(e)$height, na.rm=T)))
#this looks like it may have been entered in inches. 66 inches is 5 1/2 feet - more likely than 66 cm, which is more likely than 2.16 feet!

#hist(colData(e)$weight)
weightOutlIdx <- which(colData(e)$weight > threeSdAboveMean(mean(colData(e)$weight, na.rm=T), sd = sd(colData(e)$weight, na.rm=T)))

#if a patient is both a bmi outlier and weight (extremely above) 
#or height (extremely below) outlier, assume that they are incorrectly 
#entered in English units (not metric).
weightOutl <- intersect(weightOutlIdx, bmiOutlIdx)
heightOutl <- intersect(heightOutlIdx, bmiOutlIdx)

colData(e)[weightOutl,c("weight", "height")] #weight is likely in pounds
colData(e)[heightOutl,c("weight", "height")] #height is likely in inches

colData(e)[weightOutl,"weight"]  <- poundToKg(colData(e)[weightOutl,"weight"])
colData(e)[heightOutl,"height"]  <- inToCm(colData(e)[heightOutl,"height"])

plot(colData(e)$weight, colData(e)$height)
#recompute bmi with new values
colData(e)$bmi <- colData(e)$weight / (colData(e)$height/100)^2

```


### Normalize the clinical data

```{r}

days_last_fu <- ifelse(is.na(colData(e)$days_to_death), colData(e)$days_to_last_follow_up,colData(e)$days_to_death)
days_last_fu <- ifelse(days_last_fu < 1, 1, days_last_fu)
days_last_fu <- ifelse(days_last_fu > 365.25 * 5, 365.25 * 5, days_last_fu)

vital_stat <- ifelse(colData(e)$vital_status == "alive", 0, 1)
vital_stat <- ifelse(days_last_fu >= 365.25 * 5, 0, vital_stat)

colData(e)$days_last_fu_5yr <- days_last_fu
colData(e)$vital_stat_5yr <- vital_stat
colData(e)$bmi <- colData(e)$weight / (colData(e)$height/100)^2

#remove the patient with missing survival data
e <- e[,!is.na(colData(e)$vital_stat_5yr)]

datAddMissing <- function(colDat){
  #convert to character, add missing label
  cd <- as.character(colDat)
  cd[is.na(cd)] <- "Missing"
  return(cd)
}

colData(e)$subtype_mrna_expression_cluster <- datAddMissing(colData(e)$subtype_mrna_expression_cluster)
colData(e)$subtype_IntegrativeCluster <- datAddMissing(colData(e)$subtype_IntegrativeCluster)
colData(e)$subtype_msi_status_7_marker_call <- datAddMissing(colData(e)$subtype_msi_status_7_marker_call)
colData(e)$subtype_cna_cluster_k4 <- datAddMissing(colData(e)$subtype_cna_cluster_k4)
colData(e)$subtype_X2009stagegroup <- datAddMissing(colData(e)$subtype_X2009stagegroup)
colData(e)$subtype_histology <- datAddMissing(colData(e)$subtype_histology)
colData(e)$subtype_tumor_grade <- datAddMissing(colData(e)$subtype_tumor_grade)
colData(e)$subtype_ethnicity <- datAddMissing(colData(e)$subtype_ethnicity)
colData(e)$race <- datAddMissing(colData(e)$race)
colData(e)$seqPlate <- datAddMissing(colData(e)$seqPlate)
colData(e)$hosp <- datAddMissing(colData(e)$hosp)

subtype_mrna_mm <- model.matrix( ~ subtype_mrna_expression_cluster - 1, data=colData(e) )
subtype_integ_mm <- model.matrix( ~ subtype_IntegrativeCluster - 1, data=colData(e) )
subtype_msi_mm <- model.matrix( ~ subtype_msi_status_7_marker_call - 1, data=colData(e) )
subtype_cna_k4_mm <- model.matrix( ~ subtype_cna_cluster_k4 - 1, data=colData(e) )
stage <- model.matrix( ~ subtype_X2009stagegroup - 1, data=colData(e) )
histo <- model.matrix( ~ subtype_histology - 1, data=colData(e) )
grade <- model.matrix( ~ subtype_tumor_grade - 1, data=colData(e) )
race_mm <- model.matrix( ~ race - 1, data=colData(e) )
ethnicity_mm <- model.matrix( ~ subtype_ethnicity - 1, data=colData(e))
seqplate_mm <- model.matrix( ~ seqPlate - 1, data=colData(e))
hosp_mm <- model.matrix( ~ hosp - 1, data=colData(e))


normClinDat <- data.frame(cbind(subtype_mrna_mm, subtype_integ_mm, subtype_msi_mm, subtype_cna_k4_mm, stage, histo, grade, race_mm, ethnicity_mm, seqplate_mm, hosp_mm))
#remove redundant variables
duplicated.columns <- duplicated(t(normClinDat))
normClinDat <- normClinDat[, !duplicated.columns]

#replace continuous variables with the mean
cont_vars <- c("weight", "height", "days_to_birth", "days_last_fu_5yr", "bmi")
for(i in 1:ncol(colData(e)[,cont_vars])){
  colData(e)[,cont_vars][is.na(colData(e)[,cont_vars][,i]), i] <- median(colData(e)[,cont_vars][,i], na.rm = TRUE)
}
#replace colData with the median
#fix height issue - someone probably entered in height in feet, causing a really large BMI. Replace this by either converting to inches or replacing with the colMean (treat as missing)

rescaledDat <- data.frame(scale(colData(e)[,c("weight", "height", "days_to_birth", "bmi")]))
unscaledDat <- data.frame(colData(e)[,c("weight", "height", "days_to_birth", "bmi")])
normClinDat <- cbind(normClinDat, unscaledDat, colData(e)[,c("days_last_fu_5yr", "vital_stat_5yr")])

#remove the patient for which we have no survival information on them?
#add columns for age categories

```

## Print out the clinical data

```{r}
age <- {colData(e)$days_to_birth / -365.25} 
age %>% summary()
age %>% sd()

race <- colData(e)$race
table(race)

ethnicity <- colData(e)$ethnicity
table(ethnicity)

histo <- colData(e)$subtype_histology
table(histo)

stage <- colData(e)$subtype_X2009stagegroup
table(stage)

surv_5yr <- colData(e)$vital_stat_5yr
table(surv_5yr)

```


## Load the endometrial cancer microRNA data
```{r}

rpm_str <- "reads_per_million_miRNA_mapped_"
mir <- readRDS(file=here("data", "TCGA-UCEC-mirna.rds")) %>% 
  as_tibble %>%
  dplyr::select(miRNA_ID, starts_with("reads_per_million_miRNA_mapped")) %>%
  rename_at(vars(-miRNA_ID), funs(substr(., nchar(rpm_str)+1, nchar(.)))) %>%
  filter(rowSums(.[,2:ncol(.)]) != 0)

sum(rowSums(mir[,2:ncol(mir)]) == 0)
mir_t <- mir %>% 
  dplyr::select(which(substr(names(.), 14, 15) == "01")) %>%
  dplyr::select(which(!duplicated(substr(names(.), 1, 12)))) %>%
  dplyr::select(which(substr(names(.),1,12) %in% substr(colnames(e), 1,12)))

mir_n <- mir %>% dplyr::select(which(substr(names(.), 14, 15) == "11"))
```



## Load the nuclear hormone receptor database
```{r}
geneMap <- read_csv(here("data", "ucsc_map", "geneMap.txt")) %>% mutate(ucsc = substr(ucsc, 1,8))

nuclReceptDb <- read_csv(file=here("data", "nucl_recept_db", "targets_and_families.csv")) %>% clean_names()
## nuclReceptDb <- read.csv(url("http://www.guidetopharmacology.org/DATA/targets_and_families.csv"), header=T)
nuclReceptDb_sbst <- nuclReceptDb[nuclReceptDb$type == "nhr",]

nuclReceptDb_sbst_ensg <- nuclReceptDb_sbst %>% 
  left_join(geneMap %>% 
              dplyr::select(-ucsc, -hgnc_symbol) %>% 
              distinct, by=c("human_entrez_gene" = "entrezgene"))


```

# Identify NHRs with expression loss in a subset of tumors

The following shows the numbers for the workflow.
```{r}

filtCutoffLower <- 0.2

res <- receptLoss::receptLoss(exprMatrNml=assay(en), exprMatrTum=assay(e), nSdBelow=2, minPropPerGroup=filtCutoffLower)

#Select NHRs with rankings 1 sd above mean of all other genes. 
diffMeanSd <- mean(res %>% filter(meetsMinPropPerGrp) %>% pull(deltaMu)) + sd(res %>% filter(meetsMinPropPerGrp) %>% pull(deltaMu))

res2 <- res %>% mutate(is_nuclr_recpt = geneNm %in% nuclReceptDb_sbst_ensg$ensembl_gene_id) %>%
  left_join(nuclReceptDb_sbst_ensg, by=c("geneNm" = "ensembl_gene_id")) %>% 
  mutate(top_genes = is_nuclr_recpt & meetsMinPropPerGrp & deltaMu > diffMeanSd,
         oneSdAbove = lowerBound > 0 & meetsMinPropPerGrp & deltaMu > diffMeanSd) 

#unique(nuclReceptDb_sbst_ensg$ensembl_gene_id) %>% .[!is.na(.)] %>% length()
## Total number of:
## genes
nrow(res2)

## NHR ENSG ids
unique(nuclReceptDb_sbst_ensg$ensembl_gene_id) %>% .[!is.na(.)] %>% length()

## NHRs that overlap with genes that are expressed (selected after initial filt)
filter(res2, is_nuclr_recpt) %>% nrow()


## sum of all genes with boundary greater than 0
sum(res2$lowerBound > 0)

## sum of all nhrs with boundary greater than 0
sum({filter(res2, is_nuclr_recpt) %>% pull(lowerBound)} > 0)


## sum of all genes with boundary greater than 0 & > 20% samples per group
filter(res2, lowerBound > 0, meetsMinPropPerGrp) %>% nrow()

## sum of all nhrs with boundary greater than 0 & > 20% samples per group
filter(res2, is_nuclr_recpt, lowerBound > 0, meetsMinPropPerGrp) %>% nrow()


## number of genes (including nhrs) that are 1 sd above mean deltaMu value
filter(res2, oneSdAbove) %>% nrow()

## number of nhrs that are 1 sd above mean deltaMu value
top_nhrs <- filter(res2, is_nuclr_recpt, oneSdAbove)# %>% nrow()
nrow(top_nhrs)

## plot
set2 <- RColorBrewer::brewer.pal(8,"Set2")
set3 <- RColorBrewer::brewer.pal(12,"Set3")

delta_mu_plot <- ggplot(res2 %>% filter(meetsMinPropPerGrp, is_nuclr_recpt), aes(x=deltaMu, y = ..density..)) + 
  geom_histogram(fill=set3[5], color=set3[5], alpha=0.5, bins=40) + 
  geom_vline(xintercept = diffMeanSd, linetype='dashed') + 
  theme_classic() + 
  geom_histogram(data=res2 %>% filter(meetsMinPropPerGrp), aes(x=deltaMu, y = ..density..), 
                 fill=set3[4], color=set3[4], alpha=0.2, bins=40) + 
  ggHist + xlab("") + ylab("")
delta_mu_plot
ggsave(filename = here("output", "delta_mu_plot.png"), plot = delta_mu_plot, width = 8, height = 5, units = "in")

```


## Classify tumors into low or high expression groups using boundary

Add this classification to the `colData(e)` object

```{r}

top_nhrs_ensg <- top_nhrs$geneNm

for(i in 1:length(top_nhrs_ensg)){
  classific <- returnClassif(tum_matr = e, tum_res = res2, gene = top_nhrs$geneNm[i])
  colData(e) <- bindClassToColData(cdat = colData(e), classific)
}

```

## Check for technical bias by sequencing plate for the 6 genes.

```{r message=F, warning=F}

seqPlateClassif <- select(as.data.frame(colData(e)), starts_with("classif_")) %>%
  data.frame(seqPlate=colData(e)$seqPlate) %>%
  gather(key = "ensembl_gene_id", value = "bin_expr", 1:(ncol(.)-1))

seqPlateOrd <- orderSeqPlate(seqPlateClassif)
seqPlateCounts <- countSeqPlate(seqPlateClassif, seqPlateOrd$seqPlate)

numbRows <- 2
numbPanels <- length(unique(seqPlateCounts$ensembl_gene_id))
numbPerRow <- numbPanels / numbRows
seqPlateCounts_sbst1 <- subset(seqPlateCounts,as.character(ensembl_gene_id) %in% as.character(unique(seqPlateCounts$ensembl_gene_id)[1:numbPerRow]))
seqPlateCounts_sbst2 <- subset(seqPlateCounts,as.character(ensembl_gene_id) %in% as.character(unique(seqPlateCounts$ensembl_gene_id)[numbPerRow+ 1:numbPanels]))

ggplot(data=seqPlateCounts_sbst1, aes(x=seqPlate, y=n, fill=bin_expr)) + ggSeqPlate
ggplot(data=seqPlateCounts_sbst2, aes(x=seqPlate, y=n, fill=bin_expr)) + ggSeqPlate

#do a chi sq test between all expression plates and expression statuses
chiSqRes <- c()
for(var in unique(seqPlateCounts$ensembl_gene_id)){
  print(var)
  seqPlateCounts_sbst <- subset(seqPlateCounts, ensembl_gene_id %in% var)
  chiSqPval <- chisq.test(seqPlateCounts_sbst$bin_expr, seqPlateCounts_sbst$n)$p.value
  chiSqRes <- c(chiSqRes, chiSqPval)
}
names(chiSqRes) <- unique(seqPlateCounts$ensembl_gene_id) #levels(seqPlateCounts$ensembl_gene_id)
chiSqRes.adj <- p.adjust(chiSqRes, method="BH")
chiSqRes.adj
```

## Check for technical bias by sequencing plate for all genes.

```{r}

#boundAll <- apply(assay(en), 1, function(x) twoSdBelowMean(mean(x), sd(x)))
#boundAll is res2$lowerBound
res2_arrange <- res2 %>% arrange(geneNm)
boundAll <- res2_arrange %>% pull(lowerBound) %>% setNames(., nm = res2_arrange$geneNm)

sum(res2$lowerBound > 0) #all genes with boundary greater than 0
#is only a function of the adjacent normal data
exprLessThanBound <- rowSums(assay(e) < boundAll)
exprGrThanBound <- rowSums(assay(e) > boundAll)

filtCutoffLower <- 0.2

boundFiltGenes <- boundAll[{exprGrThanBound/ncol(assay(e)) > filtCutoffLower} & {exprGrThanBound/ncol(assay(e)) < 1-filtCutoffLower} & boundAll > 0]

binIndxMatrGrTh2 <- ifelse(assay(e)[names(boundFiltGenes),] > boundFiltGenes, 1, 0)


seqPlateClassif_all <- binIndxMatrGrTh2 %>% 
  t() %>%
  data.frame(seqPlate=colData(e)$seqPlate) %>%
  gather(key = "ensembl_gene_id", value = "bin_expr", 1:(ncol(.)-1))

seqPlateOrdAll <- orderSeqPlate(seqPlateClassif_all)
seqPlateCountsAll <- countSeqPlate(seqPlateClassif_all, seqPlateOrdAll$seqPlate)

chiSqResAll <- c()
for(var in levels(seqPlateCountsAll$seqPlate)){
  seqPlateCounts_sbst <- subset(seqPlateCountsAll, seqPlate %in% var)
  chiSqPval <- chisq.test(seqPlateCounts_sbst$bin_expr,
                          seqPlateCounts_sbst$n)$p.value
  chiSqResAll <- c(chiSqResAll, chiSqPval)
}

names(chiSqResAll) <- levels(seqPlateCountsAll$ensembl_gene_id)
chiSqResAll.adj <- p.adjust(chiSqResAll, method="BH")
min(chiSqResAll.adj)
```


# Identify survival differences based on NHR expression

```{r}

bound_nuHrRcpt_classif <- colData(e) %>% 
  as.data.frame() %>%
  select(starts_with("classif_"))
rownames(bound_nuHrRcpt_classif) <- rownames(colData(e))

abBlwDf2.nhr.top <- res2 %>% filter(is_nuclr_recpt) %>% head()

plot_surv_pval <- c()
geneNms <- c()
for(col in 1:length(colnames(bound_nuHrRcpt_classif))){
  ensg_id <- substr(x = colnames(bound_nuHrRcpt_classif)[col], 9, 23)
  gnNm <- abBlwDf2.nhr.top %>% 
    filter(ensg_id == geneNm) %>% 
    pull(hgnc_symbol)
  print(gnNm)
  print(col)
  receptLoss::plotReceptLoss(exprMatrNml=assay(en), exprMatrTum=assay(e), rldf = res2, geneName=ensg_id, clrs = set2[4:3], addToTitle = paste0(" - ", gnNm))
  
  ggsave(filename = here("output", paste0(gnNm, "_hist.png")))
  plot_surv <- plotSurvAndReturnStat(timeToEvent = colData(e)$days_last_fu_5yr/365.25, 
                                     alive0dead1 = colData(e)$vital_stat_5yr, 
                                     classif = colData(e)[,colnames(bound_nuHrRcpt_classif)[col]],
                                     surv.col = set2[1:2], gnNm = gnNm)
  ggsave(filename = here("output", paste0(gnNm, "_surv.png")), width = 5, height = 5, units = "in")
  plot_surv_pval <- c(plot_surv_pval, pvalFxn(plot_surv))
  geneNms <- c(geneNms, gnNm)
}

plot_surv_pval.adj <- p.adjust(plot_surv_pval, method = "BH")
names(plot_surv_pval.adj) <- geneNms

#expression of THRb in normal and canceroustissue 
isof = "ENSG00000151090"
tum_class <- c(colData(e)[,"classif_ENSG00000151090"], rep(2,ncol(assay(en)))) %>%
  as_tibble(x = list(cls = .))

tidyDf <- as.data.frame(cbind(as.numeric(c(assay(e)[isof,], assay(en)[isof,])),
                      as.factor(c(rep("tumor",ncol(assay(e))),
                                  rep("normal",ncol(assay(en)))))))
tidyDf %>% bind_cols(tum_class) %>% group_by(cls) %>% summarise(mn = mean(V1))

#hist of expr of PGR in normal and cancer
plotGene_bothHist(exprNml=assay(en), exprTum=assay(e), isof = "ENSG00000082175", title="", clrs = set2[4:3])
plotGene_oneHist(exprNml=assay(en), exprTum=assay(e), isof = "ENSG00000082175", title="", clrs = set2[4:3], vert_line=FALSE)

plotGene_oneHistNmlCurve(exprNml=assay(en), exprTum=assay(e), isof = "ENSG00000082175", title="", clrs = set2[4:3], vert_line=FALSE)
plotGene_oneHistNmlCurve(exprNml=assay(en), exprTum=assay(e), isof = "ENSG00000082175", title="", clrs = set2[4:3], vert_line=TRUE)
plotGene_oneHistNmlCurve(exprNml=assay(en), exprTum=assay(e), isof = "ENSG00000082175", title="", clrs = set2[4:3], vert_line=FALSE)
plotGene_oneHistNmlCurve(exprNml=assay(en), exprTum=assay(e), isof = "ENSG00000082175", title="", clrs = set2[4:3], vert_line=FALSE, histogram = FALSE)
plotGeneHist2(exprNml=assay(en), exprTum=assay(e), isof = ensg_id, title=gnNm, clrs = set2[4:3])

```

## Identify clinical features of THRB-low subtype

For the following clinical features listed in `cols_of_interest`, there is a 2 x n table showing the distribution of features between tumors that lose THRB expression (0) and tumors that do not lose THRB expression (1). Then, there are 2 additional tables: one representing the proportion of all patients within a given group, and the other representing the proportion of patients normalized to the total # of patients within each THRB group.

For each feature, following each table, there is a visual representation of the last table, containing an unadjusted p-value obtained from the chi-sq test that compares THRB expression status (0 or 1) against the variable of interest (all variables tested happen to be categorical). A list of adjusted p-values across all tested variables appears at the end of this section. The levels of each variable are represented along the outer left y axis, and their representation among each group corresponds, in order, with the colors shown within each bar plot (the inner left y axis contains an incomplete list of variables and is a plotting artifact). 

For example, for the variable "subtype_IntegrativeCluster", 'CN high' tumors represent ~5% of all tumors that lose THRB expression, while 'CN high' tumors represent ~15% of all tumors that do not lose THRB expression. The proportion of tumors represented is shown along the right y axis.

```{r}
print_tbl <- function(thrb_column, var_column){
      print(table(thrb_column, var_column) %>% prop.table %>% t %>% data.frame() %>% spread(thrb_column, Freq) %>% map_df(rev) %>% mutate(lvl_sum = `0` + `1`) %>% adorn_totals("row"))
}

print_tbl_rel <- function(thrb_column, var_column){
      print(table(thrb_column, var_column) %>% prop.table(1) %>% t %>% data.frame() %>% spread(thrb_column, Freq) %>% map_df(rev) %>% adorn_totals("row") )
}

cols_of_interest <- c("subtype_IntegrativeCluster", "subtype_msi_status_7_marker_call", "subtype_tumor_grade", "subtype_histology", "subtype_X2009stagegroup") #"subtype_cna_cluster_k4", "subtype_mrna_expression_cluster",
pval_store <- rep(NA, length(cols_of_interest))
for(i in 1:length(cols_of_interest)){
  thrb_classif <- factor(colData(e)[,"classif_ENSG00000151090"])
  col_int <- factor(colData(e)[,cols_of_interest[i]])
  pval_store[i] <- chisq.test(thrb_classif, col_int)$p.value
  print("Absolute # of patients in each group")
  print(table(colData(e)[,"classif_ENSG00000151090"], colData(e)[,cols_of_interest[i]]))
  print("Proportion of patients in each group")
  print_tbl(colData(e)[,"classif_ENSG00000151090"], colData(e)[,cols_of_interest[i]])
  print("Proportion of patients within each THRB group (0 or 1)")
  print_tbl_rel(colData(e)[,"classif_ENSG00000151090"], colData(e)[,cols_of_interest[i]])
  plot(thrb_classif, 
       col_int, 
       xlab="THRB status (0=loss of expr)",
       ylab=paste0(levels(col_int),
                   collapse = ", "),
       
       main=paste0(cols_of_interest[i], "; unadj. pval = ", round(pval_store[i],5)), 
       cex.lab=2.5, cex.axis=2.5, cex.main=1.5, cex.sub=2.5, yaxt='n', xaxt='n', ann=FALSE)
}

p.adjust(pval_store, method = "BH") %>% setNames(cols_of_interest)

## post hoc test - thrb and msi and integrative
integrative_msi_1_other_0 <- colData(e)[,cols_of_interest[1]] %>% sapply(function(x) ifelse(x == "MSI", 0, 1))
pv_integ <- stats::fisher.test(thrb_classif, integrative_msi_1_other_0) #integrative
pv_integ
msi_msih_0_other_1 <- colData(e)[,cols_of_interest[2]] %>% sapply(function(x) ifelse(x == "MSI-H", 0, 1)) #msi
pv_msi <- stats::fisher.test(thrb_classif, msi_msih_0_other_1) #integrative
pv_msi
p.adjust(c(pv_integ$p.value, pv_msi$p.value), method = "BH") %>% setNames(c("integ", "msi"))


```

## Look at survival differences between endometroid tumors
```{r}
ensg_id <- "ENSG00000151090" #THRB
gnNm <- abBlwDf2.nhr.top %>% 
  filter(ensg_id == geneNm) %>% 
  pull(hgnc_symbol)

colData_e_endometroid <- colData(e) %>% as_tibble %>% filter(subtype_histology == "Endometrioid") %>%
  DataFrame()

plot_surv <- plotSurvAndReturnStat(timeToEvent = colData_e_endometroid$days_last_fu_5yr/365.25, 
                                   alive0dead1 = colData_e_endometroid$vital_stat_5yr, 
                                   classif = colData_e_endometroid[,colnames(bound_nuHrRcpt_classif)[col]],
                                   surv.col = set2[1:2], gnNm = gnNm)

ggsave(filename = here("output", paste0(gnNm, "_endometrioid_surv.png")))
plot_surv_pval <- c(plot_surv_pval, pvalFxn(plot_surv))
```

## Look at survival differences between serous tumors
```{r}
colData_e_serous <- colData(e) %>% as_tibble %>% filter(subtype_histology == "Serous") %>%
  DataFrame()

plot_surv <- plotSurvAndReturnStat(timeToEvent = colData_e_serous$days_last_fu_5yr/365.25, 
                                   alive0dead1 = colData_e_serous$vital_stat_5yr, 
                                   classif = colData_e_serous[,colnames(bound_nuHrRcpt_classif)[col]],
                                   surv.col = set2[1:2], gnNm = gnNm)

ggsave(filename = here("output", paste0(gnNm, "_serous_surv.png")))
plot_surv_pval <- c(plot_surv_pval, pvalFxn(plot_surv))
```

# Examine combinatorial relationships between survival groups 

```{r}
colData(e)$classif_thrb_esr1_pgr <- apply(colData(e)[,c("classif_ENSG00000151090", "classif_ENSG00000091831","classif_ENSG00000082175")], 1, paste0, collapse = "") 

colData(e)$classif_esr1_pgr <- apply(colData(e)[,c( "classif_ENSG00000091831", "classif_ENSG00000082175")], 1, paste0, collapse = "") 

#must have at least 3 observations per group
subgrps_grth_3 <- colData(e)$classif_thrb_esr1_pgr %>% table %>% .[.>=3] %>% names
filt_vect <- colData(e)$classif_thrb_esr1_pgr %in% subgrps_grth_3 #c("000", "011", "111", "100")#subgrps_grth_3
plot_5combo <- plotSurvAndReturnStat(timeToEvent = {colData(e)$days_last_fu_5yr/365.25}[filt_vect], 
                            alive0dead1 = {colData(e)$vital_stat_5yr}[filt_vect], 
                            classif = {colData(e)$classif_thrb_esr1_pgr}[filt_vect],
                            surv.col = set2[c(6:4, 1:3)], gnNm = "")

#now, just look at ER/PGR negative depending on THRB status
filt_vect2 <- colData(e)$classif_thrb_esr1_pgr %in% c("000", "100")#subgrps_grth_3
filt_vect2 <- colData(e)$classif_thrb_esr1_pgr %in% c("011", "111")#subgrps_grth_3
filt_vect2 <- colData(e)$classif_thrb_esr1_pgr %in% c("100", "111")#subgrps_grth_3 #THRB pos
filt_vect2 <- colData(e)$classif_thrb_esr1_pgr %in% c("000", "011")#subgrps_grth_3 #THRB neg

plot_5combo <- plotSurvAndReturnStat(timeToEvent = {colData(e)$days_last_fu_5yr/365.25}[filt_vect2], 
                            alive0dead1 = {colData(e)$vital_stat_5yr}[filt_vect2], 
                            classif = {colData(e)$classif_thrb_esr1_pgr}[filt_vect2],
                            surv.col = set2[1:2], gnNm = "")


subgrps_esr_pgr <- colData(e)$classif_esr1_pgr %>% table %>% .[.>=3] %>% names
filt_vect3 <- colData(e)$classif_esr1_pgr %in% subgrps_esr_pgr#c("000", "011") 

plot_esr_pgr <- plotSurvAndReturnStat(timeToEvent = {colData(e)$days_last_fu_5yr/365.25}[filt_vect3], 
                            alive0dead1 = {colData(e)$vital_stat_5yr}[filt_vect3], 
                            classif = {colData(e)$classif_esr1_pgr}[filt_vect3],
                            surv.col = set2[1:3], gnNm = "")

#double positive and double negative
subgrps_esr_pgr_dp_dn <- c("00", "11")#colData(e)$classif_esr1_pgr %>% table %>% .[.>=3] %>% names
filt_vect4 <- colData(e)$classif_esr1_pgr %in% subgrps_esr_pgr_dp_dn#c("000", "011") 

plot_esr_pgr_dp_dn <- plotSurvAndReturnStat2(timeToEvent = {colData(e)$days_last_fu_5yr/365.25}[filt_vect4], 
                            alive0dead1 = {colData(e)$vital_stat_5yr}[filt_vect4], 
                            classif = {colData(e)$classif_esr1_pgr}[filt_vect4],
                            surv.col = set2[1:2], gnNm = "")

pv_00_11 <- pvalFxn(plot_esr_pgr_dp_dn$fit.stat)
pv_00_11
#plot_esr_pgr_dp_dn

#risk of death at 5 years in each group
sum_survfit_dp_dn <- summary(plot_esr_pgr_dp_dn$fit, times = 5)
sum_survfit_dp_dn$surv #risk of death at 5 years in each group
diff(sum_survfit_dp_dn$surv) #difference in 5 year risk of death btw groups


ggsave(filename = here::here("output/00_vs_11.png"), device = "png", width = 5)


#plot 000 vs 100
filt_vect5 <- colData(e)$classif_thrb_esr1_pgr %in% c("000", "100") 

# plotSurvAndReturnStat2 does the same as plotSurvAndReturnStat but just returns an additional survfit object ('fit') in a list
plot_thrb_pos_vs_neg_esr_pgr_dbl_neg <- plotSurvAndReturnStat2(timeToEvent = {colData(e)$days_last_fu_5yr/365.25}[filt_vect5], 
                            alive0dead1 = {colData(e)$vital_stat_5yr}[filt_vect5], 
                            classif = {colData(e)$classif_thrb_esr1_pgr}[filt_vect5],
                            surv.col = set2[4:5], gnNm = "")

plot_thrb_pos_vs_neg_esr_pgr_dbl_neg
pv_000_100 <- pvalFxn(plot_thrb_pos_vs_neg_esr_pgr_dbl_neg$fit.stat)
pv_000_100

p.adjust(c(pv_00_11, pv_000_100), method = "BH")


#percent alive at 5 years in each group
surv_tbl <- table({colData(e)$vital_stat_5yr}[filt_vect5], {colData(e)$classif_thrb_esr1_pgr}[filt_vect5]) %>% addmargins()
surv_perc_5yr <- surv_tbl[1,] / surv_tbl["Sum",]
surv_perc_5yr


#risk of death at 5 years in each group
sum_survfit <- summary(plot_thrb_pos_vs_neg_esr_pgr_dbl_neg$fit, times = 5)
sum_survfit$surv #risk of death at 5 years in each group
diff(sum_survfit$surv) #difference in 5 year risk of death btw groups

ggsave(filename = here::here("output/000_vs_100.png"), device = "png", width = 5)

#plot 011 vs 111

filt_vect6 <- colData(e)$classif_thrb_esr1_pgr %in% c("011", "111") 

plot_thrb_pos_vs_neg_esr_pgr_dbl_neg <- plotSurvAndReturnStat(timeToEvent = {colData(e)$days_last_fu_5yr/365.25}[filt_vect6], 
                            alive0dead1 = {colData(e)$vital_stat_5yr}[filt_vect6], 
                            classif = {colData(e)$classif_thrb_esr1_pgr}[filt_vect6],
                            surv.col = set2[1:2], gnNm = "")

#plot 100 vs 111

filt_vect7 <- colData(e)$classif_thrb_esr1_pgr %in% c("100", "111") 

plot_thrb_pos_vs_neg_esr_pgr_dbl_neg <- plotSurvAndReturnStat(timeToEvent = {colData(e)$days_last_fu_5yr/365.25}[filt_vect7], 
                            alive0dead1 = {colData(e)$vital_stat_5yr}[filt_vect7], 
                            classif = {colData(e)$classif_thrb_esr1_pgr}[filt_vect7],
                            surv.col = set2[1:2], gnNm = "")
ggsave(filename = here::here("output/100_vs_111.png"), device = "png", width = 5)

```


# Test the association between NHR expression and mutations

Read in a list of established oncogenes and tumor suppressors from [COSMIC V83](https://cancer.sanger.ac.uk/cosmic), downloaded Dec. 2, 2017. Then, select the high impact mutations as called by mutect.

```{r}
cosDat <- read.csv(file= here("data", "cosmic_mut", "Census_all_2017_12_2.csv"), header=T)

#run these two lines if readRDS does not exist
#maf_mut <- GDCquery_Maf("UCEC", pipelines = "mutect")#2018-12-09
#saveRDS(maf_mut, file = here("data", "TCGA-mut-mutect-ucec-2018-12.rds"))

mut <- readRDS(here("data","TCGA-mut-mutect-ucec-2018-12.rds"))
mut.hiImpct <- mut[mut$IMPACT %in% c("HIGH") & mut$Entrez_Gene_Id %in% cosDat$Entrez.GeneId,]
top50mutGenes <- sort(names(head(sort(table(mut.hiImpct$Hugo_Symbol),decreasing = T), 25)))
mut.hiImpct.top50 <- mut.hiImpct[mut.hiImpct$Hugo_Symbol %in% top50mutGenes,]
mut.hiImpct.top50.ptid <- names(table(mut.hiImpct.top50$Tumor_Sample_Barcode))

#which patients are associated with which gene mutations
#this is a list of patient TCGA ids
ptIdGeneMut <- sapply(top50mutGenes, function(x) unname(mut.hiImpct.top50[mut.hiImpct.top50$Hugo_Symbol %in% x,"Tumor_Sample_Barcode"]))

ptIdGeneMut.l <- lapply(ptIdGeneMut, function(x) mut.hiImpct.top50.ptid %in% x )
ptIdGeneMut.l.df <- do.call(rbind, ptIdGeneMut.l)
colnames(ptIdGeneMut.l.df) <- mut.hiImpct.top50.ptid
ptIdGeneMut.l.df <- ifelse(ptIdGeneMut.l.df, 1, 0)

#convert to summarized expt object
mutDat <- SummarizedExperiment(assays=list(mutation = as.matrix(ptIdGeneMut.l.df)))

ptInCommon <- intersect(substr(rownames(bound_nuHrRcpt_classif), 1,12), substr(colnames(assay(mutDat)), 1,12))

#create a histogram of mutation frequencies
mutFreq <- data.frame(freq=sort(rowSums(assay(mutDat[,sapply(ptInCommon, function(x) grep(x, colnames(mutDat)))])),decreasing = T))
mutFreq$gene <- rownames(mutFreq) 
mutFreq$gene <- factor(mutFreq$gene, levels=mutFreq$gene)

mf <- ggplot(mutFreq, aes(x=gene, y=freq/length(ptInCommon))) + 
  geom_bar(stat = "identity") + 
  ggBar + 
  ylab(paste0("Observed Frequency (N=", length(ptInCommon), ")"))

mf
```


## Cluster hormone receptor and mutations status

Note: Both hormone receptor expression status and mutation status are binary variables.
```{r}
#hormone receptor expression status
hrmnStatus.c <- clusterAndReorder(t(bound_nuHrRcpt_classif)[,substr(rownames(bound_nuHrRcpt_classif), 1,12) %in% ptInCommon], dimen = "both")
colnames(hrmnStatus.c) <- substr(colnames(hrmnStatus.c), 1,12)

#mutation status
mutDat.c <- clusterAndReorder(assay(mutDat)[,substr(colnames(assay(mutDat)), 1,12) %in% ptInCommon], dimen="both")
colnames(mutDat.c) <- substr(colnames(mutDat.c), 1,12)

```

## Statistical testing (Fisher's exact) between NHR expression and mutation

Are there associations between oncogene/tumor suppressor mutations and expression of NHRs? Use fisher's exact test to look at relationship between each NHR and each mutation (50x6 matrix). Our 2 matrices are mutDat.c and hrmnStatus.c.sub.

```{r}
hrmnStatus.c.sub <- hrmnStatus.c[,colnames(mutDat.c)]
rownames(hrmnStatus.c.sub) <- rownames(hrmnStatus.c.sub) %>%
                              substr(start = 9, stop = nchar(.)) %>%
                              match(abBlwDf2.nhr.top$geneNm) %>%
                              abBlwDf2.nhr.top[.,"hgnc_symbol"] %>% pull

orDat <- array(data = 0, dim = c(nrow(mutDat.c), nrow(hrmnStatus.c.sub), 2), dimnames = list(muts=rownames(mutDat.c), expr=rownames(hrmnStatus.c.sub), oddsRat=c("pValue", "OR")))

for(i in 1:nrow(mutDat.c)){
  for(j in 1:nrow(hrmnStatus.c.sub)){
    d1 <- c(mutDat.c[i,]) 
    d2 <- c(hrmnStatus.c.sub[j,])
    res <- fisher.test(x = d1, d2)
    orDat[i,j,1] <- res$p.value
    orDat[i,j,2] <- res$estimate
   }
}

p_val_adj <- orDat[,,1] %>% 
  as.vector %>% 
  p.adjust(method = "BH") %>% 
  matrix(ncol = ncol(orDat[,,1])) %>%
  as_tibble %>%
  purrr::set_names(colnames(orDat[,,1])) %>% 
  mutate(mut = rownames(orDat[,,1])) %>% 
  gather(key = "nhr", value = "val", 1:(ncol(.)-1)) %>%
  mutate(qval_or = "qval", sigStar = sig2star(val))
  
odd_rat <- orDat[,,2] %>% 
  data.frame() %>% 
  rownames_to_column("mut") %>% 
  gather(key = "nhr", value = "val", 2:ncol(.)) %>% 
  mutate(qval_or = "OR") %>% 
  as_tibble() %>%
  mutate(sigStar = "") %>%
  bind_rows(p_val_adj) %>%
  mutate(nhr = factor(nhr, levels=abBlwDf2.nhr.top$hgnc_symbol),
         mut = factor(mut, levels=rev(mutFreq$gene)))

odd_rat %>% filter(qval_or == "qval" & val < 0.05 & nhr == "THRB")

qv_signif <- odd_rat %>% filter(qval_or == "qval" & val < 0.05)
or_qv_signif <- odd_rat %>% filter(paste0(mut, nhr) %in% paste0(qv_signif$mut, qv_signif$nhr))


md <- ggplot(subset(odd_rat,qval_or=="OR"), aes(y=mut, x=nhr)) +
  geom_tile(aes(fill=log(val,10)), color="black") + 
  theme_classic() + 
  theme(text = element_text(size=20, face="italic"),
        axis.ticks=element_blank(),
        axis.title=element_blank(),
        axis.line = element_blank(), 
        axis.text.x = element_text(angle=90, hjust = 1, vjust=0.5)) +
  scale_fill_gradient2(name="log10(Odds Ratio)") +
  geom_text(data=subset(odd_rat,qval_or=="qval"), aes(label=sigStar), size=6, nudge_y = -.2) +
  coord_fixed(ratio = .65)
md

mf
#md + mf + coord_flip() + scale_y_continuous(position = "right") + scale_x_discrete(limits = rev(levels(mutFreq$gene)))

```

## Calculate Odds Ratio between each *pair* of genes

What's the likelihood that when one is expressed, the other gene is also expressed?
Use a cutoff of q = 0.01 (Fisher's Exact) for pairwise relationship. Is THRB associated with any other expression statuses?
```{r}
z <- bound_nuHrRcpt_classif
or_matrix <- matrix(ncol=ncol(z), nrow = ncol(z))
or_pval_matrix <- matrix(ncol=ncol(z), nrow = ncol(z))

for(i in 1:ncol(z)){
  for(j in 1:ncol(z)){
    if(i>=j) next
    xtab <- table(z[,i],z[,j])
    fisherTestRes <- fisher.test(xtab)
    
    oddsRat <- fisherTestRes$estimate
    or_matrix[j,i] <- oddsRat
    
    oddsRat_pval <- fisherTestRes$p.value
    or_pval_matrix[j,i] <- oddsRat_pval
  }
}

colnames(z) -> rownames(or_matrix) -> colnames(or_matrix) -> 
  rownames(or_pval_matrix) -> colnames(or_pval_matrix)

or_pval_matrix.m <- or_pval_matrix %>% 
    data.frame() %>%
  rownames_to_column("row") %>%
  gather(key = "variable", value = "pvalue", 2:ncol(.)) %>%
  filter(!is.na(pvalue)) %>%
  mutate(adjpvalue = p.adjust(pvalue, method="BH"),
         stars = cut(adjpvalue, 
                     breaks=c(-Inf, 0.0001, 0.001, 0.01, Inf), 
                     label=c("***", "**", "*", ""))) %>%
  mutate(row_var = paste0(row,":",variable)) %>%
  dplyr::select(-row, -variable)

or_matrix.m <- or_matrix %>%
  data.frame() %>%
  rownames_to_column("row") %>%
  gather(key = "variable", value = "OR", 2:ncol(.)) %>%
  filter(!is.na(OR)) %>%
  mutate(OR_cut = cut(OR ,breaks=c(0, 0.01, 0.1, 0.5, 2, 5, 50, Inf),
                      include.lowest=TRUE,
                      label=c("(0,0.01)","(0.01,0.1)", "(0.1,0.5)", 
                              "(0.5,2)", "(2,5)", "(5,50)", "(50,Inf)"))) %>%
  mutate(row_var = paste0(row,":",variable)) %>%
  dplyr::select(-row, -variable) %>% 
  left_join(or_pval_matrix.m, by="row_var") %>%
  separate(row_var, into = c("col1", "col2"), sep = ":")

gene_key <- data.frame(geneNms)
rownames(gene_key) <- colnames(bound_nuHrRcpt_classif)

or_matrix.m2 <- or_matrix.m %>% 
  mutate(col1 = factor(gene_key[col1,], levels=gene_key$geneNms),
         col2=factor(gene_key[col2,], levels=gene_key$geneNms))

colors_orr_corr_matr <- append(brewer.pal(11,"PuOr")[c(2,4,5,7,8,10)], "#ffffff", after=3) 

fig3a2 <- ggplot(or_matrix.m2, aes(col1, col2)) + 
  geom_tile(aes(fill=OR_cut),colour="black", size=1.5) + 
  geom_text(aes(label=stars), color="black", size=8, vjust=1) + 
  theme_classic() + 
  theme(axis.line = element_blank(), 
        text = element_text(size=20), 
        axis.text =element_text(size=16, color="black", face="italic"), 
        axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5),
        axis.ticks = element_blank(), 
        legend.key.size = unit(1.25,"cm")) + 
  scale_fill_manual(values = rev(colors_orr_corr_matr), 
                    name="Odds Ratio", drop=FALSE) + 
  coord_fixed(ratio = 1) + 
  scale_y_discrete(position="right") + 
  xlab("") + ylab("")

fig3a2

or_matrix.m2
```

# Differential microRNA expression analysis 

We look at the association between miRNAs and THRB expression

## Volcano plot showing miRNA expression
```{r}

bound_nuHrRcpt_classif_mirna_filt <- bound_nuHrRcpt_classif %>% .[rownames(.) %>% substr(1,12) %in% substr(colnames(mir_t), 1,12),]


mir_t2 <- rownames(bound_nuHrRcpt_classif_mirna_filt) %>% 
  substr(1,12) %>%
  sapply(grep, colnames(mir_t)) %>%
  unlist %>%
  unique %>%
  mir_t[,.] %>%
  mutate_all(.funs = ~log(.+1, 2))

mir_n2 <- colnames(en) %>% 
  substr(1,12) %>%
  sapply(grep, colnames(mir_n)) %>%
  unlist %>%
  mir_n[,.] %>%
  mutate_all(.funs = ~log(.+1, 2))

e_sub <- substr(colnames(mir_t2), 1,12) %>% 
  sapply(grep,colnames(e)) %>%
  e[,.]
en_sub <- substr(colnames(mir_n2), 1,12) %>% 
  sapply(grep,colnames(en)) %>%
  en[,.]


#############
i=6
classif <- colnames(bound_nuHrRcpt_classif)[i]#"classif_ENSG00000082175"
vars <- c(classif, "seqPlate", "hosp")
colData_e_stdze <- colData(e_sub)[,vars]
dsgn <- model.matrix(~ colData(e_sub)[,classif] + colData(e_sub)[,"seqPlate"] + colData(e_sub)$hosp)
allDeGenes <- returnDeMirnaLimma(mir_t2, dsgn, colOfInt = 2, retAll=T)

logFCThr = 1
adjPvalThresh = 0.001

allDeGenes2 <- allDeGenes %>% rownames() %>% 
  as.numeric() %>% mir$miRNA_ID[.] %>%
  cbind(allDeGenes, mirna=.) %>%
  as_tibble() %>%
  mutate(mirna = as.character(mirna),
         mirna_label = ifelse(abs(logFC) > logFCThr & adj.P.Val < adjPvalThresh,
                              mirna,""))

rowIdxDeMirna <- rownames(allDeGenes)[1] %>% as.numeric
mir$miRNA_ID[rowIdxDeMirna]
#make a volcano plot
ggplot(allDeGenes2, 
       aes(x = logFC, y = -log(adj.P.Val, 10))) +
  theme_bw() + geom_vline(xintercept = c(-logFCThr,logFCThr), 
                          linetype="dashed", color = "grey") + 
  geom_hline(yintercept = -log(adjPvalThresh,10), 
             linetype="dashed", color = "grey") +
  geom_point(alpha=0.5)+ 
  geom_text_repel(aes(label=mirna_label))+#aes(color=!inNeither & numHits > 3)) + 
  theme_classic() +
  theme(text  = element_text(size=18), legend.title = element_blank()) + 
  labs(size = 14, x = expression(paste(log[2], "(miRNA in THRB_Hi/THRB_Low)")), 
       y = (expression(-log[10](q~value)))) + 
  #annotate("segment", x=-0.7, y=0, xend=-1.15, yend=1.9, size=2, arrow=arrow(length = unit(0.03, "npc"), type="closed")) + 
  annotate("segment", x=-1.3, y=2.6, xend=-1.2, yend=3.5, size=1) +#, arrow=arrow(length = unit(0.03, "npc"), type="closed")) + 
  
  #annotate("text", x=-1.5, y=1, label="Upregulated in\nendomet. tumors\nwith low THRB expr.", hjust = 0) + 
  annotate("rect", xmin=-Inf, xmax=-1, ymin=3, ymax=Inf, alpha=0.2, fill="red") +
  annotate("text", x=-1.5, y=2, label="miRNAs upregulated\nin endomet. tumors\nwith low THRB expr.", hjust = 0)


#add arrow

```

## Volcano plot showing miRNA differential expression in tumors with endometrioid histology


```{r}

colData(e)$subtype_histology #will do differential expression for the endometrioid and serous tumors
# Endometrioid      Missing        Mixed       Serous 
#          304          172           13           53

ec_histology <- "Endometrioid"#"Serous" #

tcga_id_with_histo <- colData(e)[colData(e)$subtype_histology == ec_histology,] %>% rownames()

bound_nuHrRcpt_classif_mirna_filt2 <- bound_nuHrRcpt_classif %>% .[rownames(.) %>% substr(1,12) %in% substr(colnames(mir_t), 1,12),] %>% .[rownames(.) %>% substr(1,12) %in% substr(tcga_id_with_histo, 1,12),] 

mir_t3 <- rownames(bound_nuHrRcpt_classif_mirna_filt2) %>% 
  substr(1,12) %>%
  sapply(grep, colnames(mir_t)) %>%
  unlist %>%
  unique %>%
  mir_t[,.] %>%
  mutate_all(.funs = ~log(.+1, 2))

mir_n3 <- colnames(en) %>% 
  substr(1,12) %>%
  sapply(grep, colnames(mir_n)) %>%
  unlist %>%
  mir_n[,.] %>%
  mutate_all(.funs = ~log(.+1, 2))

e_sub <- substr(colnames(mir_t3), 1,12) %>% 
  sapply(grep,colnames(e)) %>%
  e[,.]
en_sub <- substr(colnames(mir_n3), 1,12) %>% 
  sapply(grep,colnames(en)) %>%
  en[,.]


#############
i=6
classif <- colnames(bound_nuHrRcpt_classif)[i]#"classif_ENSG00000082175"
vars <- c(classif, "seqPlate", "hosp")
colData_e_stdze <- colData(e_sub)[,vars]
dsgn <- model.matrix(~ colData(e_sub)[,classif] + colData(e_sub)[,"seqPlate"] + colData(e_sub)$hosp)
allDeGenes <- returnDeMirnaLimma(mir_t3, dsgn, colOfInt = 2, retAll=T)

logFCThr = 1
adjPvalThresh = 0.001

allDeGenes2 <- allDeGenes %>% rownames() %>% 
  as.numeric() %>% mir$miRNA_ID[.] %>%
  cbind(allDeGenes, mirna=.) %>%
  as_tibble() %>%
  mutate(mirna = as.character(mirna),
         mirna_label = ifelse(abs(logFC) > logFCThr & adj.P.Val < adjPvalThresh,
                              mirna,""))

rowIdxDeMirna <- rownames(allDeGenes)[1] %>% as.numeric
mir$miRNA_ID[rowIdxDeMirna]
#make a volcano plot
ggplot(allDeGenes2, 
       aes(x = logFC, y = -log(adj.P.Val, 10))) +
  theme_bw() + geom_vline(xintercept = c(-logFCThr,logFCThr), 
                          linetype="dashed", color = "grey") + 
  geom_hline(yintercept = -log(adjPvalThresh,10), 
             linetype="dashed", color = "grey") +
  geom_point(alpha=0.5)+ 
  geom_text_repel(aes(label=mirna_label))+
  theme_classic() +
  theme(text  = element_text(size=18), legend.title = element_blank()) + 
  labs(size = 14, x = expression(paste(log[2], "(miRNA in THRB_Hi/THRB_Low)")), 
       y = (expression(-log[10](q~value)))) + 
  annotate("segment", x=-1.3, y=2.6, xend=-1.2, yend=3.5, size=1) +
  
  annotate("rect", xmin=-Inf, xmax=-1, ymin=3, ymax=Inf, alpha=0.2, fill="red") +
  annotate("text", x=-1.5, y=2, label="miRNAs upregulated\nin endomet. tumors\nwith low THRB expr.", hjust = 0) +
    ggtitle("Diff miRNA expr in endometrial\ntumors with endometrioid histology")



```


## Volcano plot showing miRNA differential expression in tumors with serous histology

```{r}

ec_histology <- "Serous"

tcga_id_with_histo <- colData(e)[colData(e)$subtype_histology == ec_histology,] %>% rownames()

bound_nuHrRcpt_classif_mirna_filt2 <- bound_nuHrRcpt_classif %>% .[rownames(.) %>% substr(1,12) %in% substr(colnames(mir_t), 1,12),] %>% .[rownames(.) %>% substr(1,12) %in% substr(tcga_id_with_histo, 1,12),] 

mir_t3 <- rownames(bound_nuHrRcpt_classif_mirna_filt2) %>% 
  substr(1,12) %>%
  sapply(grep, colnames(mir_t)) %>%
  unlist %>%
  unique %>%
  mir_t[,.] %>%
  mutate_all(.funs = ~log(.+1, 2))

mir_n3 <- colnames(en) %>% 
  substr(1,12) %>%
  sapply(grep, colnames(mir_n)) %>%
  unlist %>%
  mir_n[,.] %>%
  mutate_all(.funs = ~log(.+1, 2))

e_sub <- substr(colnames(mir_t3), 1,12) %>% 
  sapply(grep,colnames(e)) %>%
  e[,.]
en_sub <- substr(colnames(mir_n3), 1,12) %>% 
  sapply(grep,colnames(en)) %>%
  en[,.]


#############
i=6
classif <- colnames(bound_nuHrRcpt_classif)[i]#"classif_ENSG00000082175"
vars <- c(classif, "seqPlate", "hosp")
colData_e_stdze <- colData(e_sub)[,vars]
dsgn <- model.matrix(~ colData(e_sub)[,classif] + colData(e_sub)[,"seqPlate"] + colData(e_sub)$hosp)
allDeGenes <- returnDeMirnaLimma(mir_t3, dsgn, colOfInt = 2, retAll=T)

logFCThr = 1
adjPvalThresh = 0.001

allDeGenes2 <- allDeGenes %>% rownames() %>% 
  as.numeric() %>% mir$miRNA_ID[.] %>%
  cbind(allDeGenes, mirna=.) %>%
  as_tibble() %>%
  mutate(mirna = as.character(mirna),
         mirna_label = ifelse(abs(logFC) > logFCThr & adj.P.Val < adjPvalThresh,
                              mirna,""))

rowIdxDeMirna <- rownames(allDeGenes)[1] %>% as.numeric
mir$miRNA_ID[rowIdxDeMirna]
#make a volcano plot
ggplot(allDeGenes2, 
       aes(x = logFC, y = -log(adj.P.Val, 10))) +
  theme_bw() + geom_vline(xintercept = c(-logFCThr,logFCThr), 
                          linetype="dashed", color = "grey") + 
  geom_hline(yintercept = -log(adjPvalThresh,10), 
             linetype="dashed", color = "grey") +
  geom_point(alpha=0.5)+ 
  geom_text_repel(aes(label=mirna_label))+#aes(color=!inNeither & numHits > 3)) + 
  theme_classic() +
  theme(text  = element_text(size=18), legend.title = element_blank()) + 
  labs(size = 14, x = expression(paste(log[2], "(miRNA in THRB_Hi/THRB_Low)")), 
       y = (expression(-log[10](q~value)))) + 
  #annotate("segment", x=-0.7, y=0, xend=-1.15, yend=1.9, size=2, arrow=arrow(length = unit(0.03, "npc"), type="closed")) + 
  annotate("segment", x=-1.3, y=2.6, xend=-1.2, yend=3.5, size=1) +#, arrow=arrow(length = unit(0.03, "npc"), type="closed")) + 
  
  #annotate("text", x=-1.5, y=1, label="Upregulated in\nendomet. tumors\nwith low THRB expr.", hjust = 0) + 
  annotate("rect", xmin=-Inf, xmax=-1, ymin=3, ymax=Inf, alpha=0.2, fill="red") +
  annotate("text", x=-1.5, y=2, label="miRNAs upregulated\nin endomet. tumors\nwith low THRB expr.", hjust = 0) + 
  ggtitle("Diff miRNA expr in endometrial\ntumors with serous histology")


```



## Expression of miRNA-146a across various tissue types

```{r}
mir146a_expr_adj_nml <- as_tibble(list(id = 2, mir_expr = as.numeric(mir_n2[rowIdxDeMirna,])))

mir146a_expr <- as_tibble(list(id = bound_nuHrRcpt_classif_mirna_filt[,6], 
                          mir_expr = as.numeric(mir_t2[rowIdxDeMirna,]))) %>%
  bind_rows(mir146a_expr_adj_nml) %>%
  mutate(id = factor(id),
         id = recode(id, "0" = "Loss", #loss = THRB-low
         "1" = "Baseline", #THRB-high
         "2" = "Adj. Normal")) 

#statistical test to check difference between 2 groups
grp0 <- mir146a_expr %>% filter(as.character(id) == "Loss") %>% pull(mir_expr) %>% as.numeric()
grp1 <- mir146a_expr %>% filter(as.character(id) == "Baseline") %>% pull(mir_expr) %>% as.numeric()
grp2 <- mir146a_expr %>% filter(as.character(id) == "Adj. Normal") %>% pull(mir_expr) %>% as.numeric()

p1 <- wilcox.test(grp0, grp1) #p1$p.value 
p2 <- wilcox.test(grp1, grp2) 
p3 <- wilcox.test(grp0, grp2) 

p.adjust(c(p1$p.value, p2$p.value, p3$p.value), method = "BH")


ggplot(mir146a_expr, aes(x=id, y=mir_expr, fill=id)) + 
  geom_violin() + 
  geom_jitter(pch=21, alpha=0.5) + 
  ylab("miRNA-146a expression") +
  xlab("") + 
  #ggtitle("Expression of miRNA-146a\nin endometrial cancer") + 
  scale_fill_manual(values=set2[c(1,2,4)]) +
  theme_classic() + 
  theme(legend.position="none", text=element_text(size=18, color="black"))
  
ggsave(filename = here::here("output/mir146.png"), width = 5.5, height = 4, units = "in")
```


## Are the THRB-low tumors the same as the Type I endometrial cancers?


```{r}
colDat_e_subst <- colData(e) %>% as_tibble %>% filter(subtype_histology == "Endometrioid" | subtype_histology == "Serous")#, Serous"))
#colData(e)$classif_ENSG00000151090
#varsToTest <- colData(e) %>% as_tibble %>% dplyr::select(patient, starts_with("subtype"))
ft <- fisher.test(colDat_e_subst$subtype_histology, colDat_e_subst$classif_ENSG00000151090)
table(colDat_e_subst$subtype_histology, colDat_e_subst$classif_ENSG00000151090)
ft
```

The THRB-low tumors (with better survival) are more likely to be endometrioid (type I), pvalue = `r round(ft$p.value, 3)`, OR = `r round(ft$estimate, 3)`.


# THRB expression in other cancers

we look at THRB expression patterns in other cancer types (in both tumor and normal tissue) and ask whether THRB expression is predictive of survival in other cancers.

```{r}

brca <- readRDS(here("data", "TCGA-BRCA-exp.rds")) %>% clean_dataset()
luad <- readRDS(here("data", "TCGA-LUAD-exp.rds")) %>% clean_dataset()
lusc <- readRDS(here("data", "TCGA-LUSC-exp.rds")) %>% clean_dataset()

brca_n <- brca$en
brca <- brca$et

luad_n <- luad$en
luad <- luad$et

lusc_n <- lusc$en
lusc <- lusc$et

ucec_n <- en
ucec <- e

names_of_dfs <- c("ucec_n","brca_n", "luad_n", "lusc_n", "brca", "luad", "lusc", "ucec")

esr1 = "ENSG00000091831"
ar = "ENSG00000169083"
pr = "ENSG00000082175"
thrb = "ENSG00000151090"
thra = "ENSG00000126351"
recepts = c(esr1, ar, pr, thrb, thra)
```

## What does THRB receptor expression look like in other cancers?

```{r}

thrb_df <- lapply(names_of_dfs, function(x) assay_to_df(eval(parse(text = x)), tum_type = x, gene = thrb)) %>%
  do.call(rbind, .) %>% 
  mutate(class_temp = substr(tum_type, start = nchar(tum_type)-1, stop=nchar(tum_type)),
                   class = ifelse(class_temp == "_n", "adj_nml", "tumor")) %>%
  select(-class_temp) %>%
  mutate(tum_type2 = substr(tum_type, 1,4),
         class=factor(class),
         tum_name = factor(tum_type2, levels = c("ucec", "brca", "luad", "lusc"),
                           labels = c("Uterus", "Breast", "Lung Adeno", "Lung Squam")))


ggplot(thrb_df, aes(x=tum_name, y=ENSG00000151090, color=class)) +
  scale_color_manual(values=set2[4:3]) + 
  geom_boxplot(fill=NA, outlier.shape = NA, size=1, alpha=0.8) + 
  theme_classic() + 
  geom_point(pch = 21, alpha=0.8,
             position = position_jitterdodge(jitter.height = 0, jitter.width = 0.5)) + 
  xlab("") + ylab("THRB expression (log2(TPM))")

```

## Run receptLoss on other cancer types and look at THRB expression here

```{r}
brca_res <- receptLoss(exprMatrNml=assay(brca_n), exprMatrTum=assay(brca), nSdBelow=2, minPropPerGroup=0.2)
luad_res <- receptLoss(exprMatrNml=assay(luad_n), exprMatrTum=assay(luad), nSdBelow=2, minPropPerGroup=0.2)
lusc_res <- receptLoss(exprMatrNml=assay(lusc_n), exprMatrTum=assay(lusc), nSdBelow=2, minPropPerGroup=0.2)

receptLoss::plotReceptLoss(exprMatrNml = assay(brca_n), exprMatrTum = assay(brca), rldf = brca_res, geneName = thrb, clrs = set2[4:3])

receptLoss::plotReceptLoss(exprMatrNml = assay(luad_n), exprMatrTum = assay(luad), rldf = luad_res, geneName = thrb, clrs = set2[4:3])

receptLoss::plotReceptLoss(exprMatrNml = assay(lusc_n), exprMatrTum = assay(lusc), rldf = lusc_res, geneName = thrb, clrs = set2[4:3])
```

## Is there a difference in survival between high and low groups based on THRB expression in other tumors? Classify the tumors first, then perform K-M survival analysis

```{r}

colData(brca) <- bindClassToColData(cdat = colData(brca), returnClassif(brca, brca_res, thrb))  %>%
  bindClassToColData(obj = calc_days_last_fu(colData(brca)$days_to_death,
                                             colData(brca)$days_to_last_follow_up)) %>% 
  bindClassToColData(calc_vital_stat(vital_stat = colData(brca)$vital_status, .$days_last_fu))

colData(luad) <- bindClassToColData(cdat = colData(luad), 
                                    obj = returnClassif(luad, luad_res, thrb)) %>%
  bindClassToColData(obj = calc_days_last_fu(colData(luad)$days_to_death,
                                             colData(luad)$days_to_last_follow_up)) %>% 
  bindClassToColData(calc_vital_stat(vital_stat = colData(luad)$vital_status, .$days_last_fu))

colData(lusc) <- bindClassToColData(cdat = colData(lusc), 
                                    obj = returnClassif(lusc, lusc_res, thrb)) %>%
  bindClassToColData(obj = calc_days_last_fu(colData(lusc)$days_to_death,
                                             colData(lusc)$days_to_last_follow_up)) %>% 
  bindClassToColData(calc_vital_stat(vital_stat = colData(lusc)$vital_status, .$days_last_fu))


plotSurvAndReturnStat(timeToEvent = colData(brca)$days_last_fu/365.25, alive0dead1 = colData(brca)$vital_stat, classif = colData(brca)$classif_ENSG00000151090, surv.col = set2[1:2], gnNm = "BRCA survival based on THRB expression")

plotSurvAndReturnStat(timeToEvent = colData(luad)$days_last_fu/365.25, alive0dead1 = colData(luad)$vital_stat, classif = colData(luad)$classif_ENSG00000151090, surv.col = set2[1:2], gnNm = "LUAD survival based on THRB expression")

plotSurvAndReturnStat(timeToEvent = colData(lusc)$days_last_fu/365.25, alive0dead1 = colData(lusc)$vital_stat, classif = colData(lusc)$classif_ENSG00000151090, surv.col = set2[1:2], gnNm = "LUSC survival based on THRB expression")


```


# Session Information

```{r}
sessionInfo()
```

## Supplementary scripts

The following scripts are found in the file `2019_endomet_smry_suppl.R`.

```{r eval=F}

orderSeqPlate <- function(x){
  x %>% 
    group_by(seqPlate) %>%
    dplyr::count(seqPlate) %>% 
    arrange(-n)
}

countSeqPlate <- function(x, sqPlates){
  x %>%
    group_by(bin_expr, seqPlate, ensembl_gene_id) %>%
    dplyr::count(seqPlate) %>%
    ungroup() %>%
    mutate(seqPlate = factor(seqPlate, levels=sqPlates)) %>%
    mutate(bin_expr = factor(bin_expr, c(1,0)))
}

#scripts for doing GSEA with Fisher's exact test
#R package: gSEAfish - Set enrichment analysis using Fisher's exact test
#also provides visualization tools
 
genesetEnrichmentFisher <- function(B, Targ, GS.GS, GS.names){ 
  #calculates the fisher exact t test for enrichment in a particular set of
  #target gene entrez IDs (T) among the number of background genes (B). GS.GS
  #takes the form "hGMT$genesets", where hGMT is a variable that stores the .gmt
  #file object downloaded from the msigdb website that was read in using the
  #GMT() function. GS.names takes the form "hGMT$geneset.names"
  
  #1. Initialize an empty dataframe
  t <- unlist(fisher.test(matrix(data = c(1,4,6,2), nrow = 2), conf.int = T))
  colNam <- c(names(t), "geneset.names", "MatchesBtwTargAndPathway", "NumbGenesInMsigdbPathway", "TargLen", "BackgroundLen")
  fisherDf <- data.frame(matrix(NA, nrow=1, ncol=13))
  names(fisherDf) <- colNam
  
  TargLen <- length(Targ)
  #2. Loop through each geneset and calculate pvalues
  for(i in 1:length(GS.GS)){
    pathwayEntrezIDs <- GS.GS[[i]] #these are characters
    M <- length(intersect(Targ, pathwayEntrezIDs)) #number of overlapping genes
    P <- length(pathwayEntrezIDs) #numb of genes in pathway
    conting.matrix <- matrix(c(M, TargLen-M, P-M, B-TargLen-P+M), nrow = 2, byrow = TRUE) #create contigency matrix
    fisherTest <- fisher.test(conting.matrix, conf.int = T)
    fisherDf <- rbind(fisherDf, c(unlist(fisherTest), GS.names[[i]], M, P, TargLen, B))
  }
  fisherDf = fisherDf[-1,]   #remove the row of NA's (in row 1)
  fisherDf$bhPval <- p.adjust(fisherDf$p.value, method = "BH")
  fisherDf$bonferroni <- p.adjust(fisherDf$p.value, method = "bonferroni")
  fisherDf$neg.log10.bhPval <- -log10(as.numeric(fisherDf$bhPval))
  fisherDf$dataset <-
    unlist(lapply(strsplit(fisherDf$geneset.names, "_"), function(x) x[1]))
  fisherDf <- fisherDf[order(fisherDf$bhPval),]
  #rank by bhPval pbal
  return(fisherDf)
}


forestplot <- function(d, title, xlab=expression(log[10] *"(Odds Ratio)"), ylab="Pathway", size=12){
  p <- ggplot(d, aes(x=x, y=y, ymin=ylo, ymax=yhi)) +
    geom_pointrange(size = 1.25) +
    coord_flip() +
    #geom_hline(aes(x=0, yintercept = 1)) +
    ylab(xlab) +
    xlab(ylab) + #switch because of the coord_flip() above
    ggtitle(title) +
    scale_y_log10() +
    theme_classic() +
    geom_text(aes(label=numbMatches)) +
    guides(alpha=FALSE) + 
    theme(axis.text.x=element_text(size=size), axis.text.y = element_text(size = 7), legend.text = element_text(size = 12))
  return(p)
}

createForestDF <- function(data, title){
  d <- data.frame(x=data$geneset.names, y=as.numeric(data$`estimate.odds ratio`), ylo = as.numeric(data$conf.int1), yhi = as.numeric(data$conf.int2), pval=as.numeric(data$p.value), numbGeneInPway = data$NumbGenesInMsigdbPathway,
                  numbMatches=data$MatchesBtwTargAndPathway)
  d$x <-factor(d$x, levels=d[order(-d$y), "x"])
  return(forestplot(d, title, xlab=expression("Odds Ratio("*log[10] *" scale)")))
}

gseWrapper <- function(B, Targ, genesets, geneset.names, qval, OR){
  #dir can equal "up" or "down" for up or down regulated, respectivtly
  res_gse <- genesetEnrichmentFisher(B = B, Targ = Targ, GS.GS = genesets, GS.names=geneset.names)
  res_gse_filt <- res_gse[as.numeric(res_gse$bhPval) < qval & as.numeric(res_gse$`estimate.odds ratio`) > OR & as.numeric(res_gse$conf.int1) > OR,]
  return(res_gse_filt)
}


readHmKgRctmepathways <- function(gsea_path){
  hallmark <- GSA.read.gmt(paste0(gsea_path,"h.all.v5.1.entrez.gmt")) #50 genesets
  kegg <- GSA.read.gmt(paste0(gsea_path,"c2.cp.kegg.v5.1.entrez.gmt")) #185 genesets
  reactome <- GSA.read.gmt(paste0(gsea_path,"c2.cp.reactome.v5.1.entrez.gmt")) #673 genesets
  hallmarkKeggReactome <- list(genesets = c(hallmark$genesets, kegg$genesets, reactome$genesets), geneset.names =
                                 c(hallmark$geneset.names, kegg$geneset.names, reactome$geneset.names), geneset.descriptions =
                                 c(hallmark$geneset.descriptions, kegg$geneset.descriptions, reactome$geneset.descriptions))
  return(hallmarkKeggReactome)
}

ensgToEntrez2 <- function(ensg, geneMap){
  g <- geneMap %>% 
    dplyr::select(-ucsc) %>% 
    distinct %>% 
    filter(ensembl_gene_id %in% ensg)
  return(g)
}


geneSymbToEnsg <- function(geneSymb){
  ensembl = useMart("ensembl",dataset="hsapiens_gene_ensembl")
  geneEntrez <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "entrezgene"),
                      filters = "hgnc_symbol",
                      values = geneSymb,
                      mart = ensembl)
  return(geneEntrez)
}






twoSdBelowMean <- function(mn, sd){
  twosdblw <- mn - 2*sd
  return(twosdblw)
}

twoSdAboveMean <- function(mn, sd){
  twosdabv <- mn + 2*sd
  return(twosdabv)
}

threeSdBelowMean <- function(mn, sd){
  threesdblw <- mn - 3*sd
  return(threesdblw)
}

threeSdAboveMean <- function(mn, sd){
  threesdabv <- mn + 3*sd
  return(threesdabv)
}
fourSdBelowMean <- function(mn, sd){
  fsdblw <- mn - 4*sd
  return(fsdblw)
}

fourSdAboveMean <- function(mn, sd){
  fsdabv <- mn + 4*sd
  return(fsdabv)
}

poundToKg <- function(lb){
  kg_per_lb = 2.2046226218
  return(lb/kg_per_lb)
}

inToCm <- function(inches){
  cm_per_in = 2.54
  return(cm_per_in*inches)
}




plotSurvAndReturnStat <- function(timeToEvent, alive0dead1, classif, surv.col, gnNm=""){
  fit <- survfit(Surv(timeToEvent, alive0dead1) ~ classif)
  fit.stat <- survdiff(Surv(timeToEvent, alive0dead1) ~ classif)
  print(fit.stat)
  #650x250
  p <- GGally::ggsurv(fit, back.white = T, surv.col = surv.col, size.est = 1.25) + 
    ylim(c(0,1)) + theme_classic() + 
    ggtitle(gnNm) +
    theme(aspect.ratio = 1, legend.title=element_blank(), 
          text = element_text(size=42, color = "black"), 
          axis.title=element_blank(),
          plot.title =element_text(size = 54, margin=margin(0,0,-5,0), face="italic")) +
  #add the number of patients
  annotate("text", x = 1, y = .35, label = paste0(names(fit.stat$n)[1], ", N = ",fit.stat$n[1])) +
  annotate("text", x = 1, y = .15, label = paste0(names(fit.stat$n)[2], ", N = ",fit.stat$n[2])) +
  theme(plot.margin=grid::unit(c(0,0,0,0), "mm"))

  print(p)
  return(fit.stat)
}

plotSurvAndReturnStat2 <- function(timeToEvent, alive0dead1, classif, surv.col, gnNm=""){
  fit <- survfit(Surv(timeToEvent, alive0dead1) ~ classif)
  fit.stat <- survdiff(Surv(timeToEvent, alive0dead1) ~ classif)
  print(fit.stat)
  #650x250
  p <- GGally::ggsurv(fit, back.white = T, surv.col = surv.col, size.est = 1.25) + 
    ylim(c(0,1)) + theme_classic() + 
    ggtitle(gnNm) +
    theme(aspect.ratio = 1, legend.title=element_blank(), 
          text = element_text(size=25, color = "black"), 
          axis.title=element_blank(),
          plot.title =element_text(face="italic")) +
    #add the number of patients
    annotate("text", x = 1, y = .35, label = paste0(names(fit.stat$n)[1], ", N = ",fit.stat$n[1])) +
    annotate("text", x = 1, y = .15, label = paste0(names(fit.stat$n)[2], ", N = ",fit.stat$n[2])) +
    theme(plot.margin=grid::unit(c(0,0,0,0), "mm"))
  
  print(p)
  return(list(fit.stat = fit.stat, fit = fit))
}

plotSurvAndReturnStat2_likrat <- function(timeToEvent, alive0dead1, classif, surv.col, gnNm=""){
  fit <- survfit(Surv(timeToEvent, alive0dead1) ~ classif)
  fit.stat <- survdiff(Surv(timeToEvent, alive0dead1) ~ classif)
  print(fit.stat)
  #650x250
  p <- GGally::ggsurv(fit, back.white = T, surv.col = surv.col, size.est = 1.25) + 
    ylim(c(0,1)) + theme_classic() + 
    ggtitle(gnNm) +
    theme(aspect.ratio = 1, legend.title=element_blank(), 
          text = element_text(size=25, color = "black"), 
          axis.title=element_blank(),
          plot.title =element_text(face="italic")) +
    #add the number of patients
    annotate("text", x = 1, y = .35, label = paste0(names(fit.stat$n)[1], ", N = ",fit.stat$n[1])) +
    annotate("text", x = 1, y = .15, label = paste0(names(fit.stat$n)[2], ", N = ",fit.stat$n[2])) +
    theme(plot.margin=grid::unit(c(0,0,0,0), "mm"))
  
  print(p)
  return(list(fit.stat = fit.stat, fit = fit))
}


returnDeGenesLimma <- function(expr, design, colOfInt, pval=1, logFC=0, retAll=F){
  fit_t <- lmFit(object = assay(expr), design = design)
  ebfit_t <- eBayes(fit_t)
  if(retAll == T){
    topTableRes_t <- topTable(ebfit_t, coef = colOfInt, number = Inf) 
    return(topTableRes_t)
  }
  
  topTableRes_t <- topTable(ebfit_t, coef = colOfInt, number = nrow(assay(expr)), adjust.method = "BH", p.value = pval, lfc = logFC) 
  return(topTableRes_t)
}

bindClassToColData <- function(cdat, obj){
  #if the columns exist, replace them
  #if the column does not exist, add it
  cno <- colnames(obj)
  for(col in 1:ncol(obj)){
    cdat$temp1 <- obj[,col] #add a temporary column from obj
    if(cno[col] %in% colnames(cdat)){
      #the column does exist, so replace it
      cdat[, cno[col]] <- NULL
    }
    colnames(cdat)[which(colnames(cdat) == "temp1")] <- cno[col]
  }
  return(cdat)
}


plotGeneHist2 <- function(exprNml, exprTum, isof, title, clrs){

  tidyDf <- as.data.frame(cbind(as.numeric(c(exprTum[isof,], exprNml[isof,])),
                                as.factor(c(rep("tumor",ncol(exprTum)), rep("normal",ncol(exprNml))))),
                          stringsAsFactors=FALSE)
  colnames(tidyDf) <- c("expr", "type")
  print(mean(exprNml[isof,]))
  print(twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
  expr <- type <- ..density.. <- NULL # Setting the variables to NULL first
  p1 <- ggplot(tidyDf, aes(x=expr, color=as.factor(type),
                           fill=as.factor(type),
                           group=as.factor(type))) +
    theme_classic() +
    geom_histogram(data=subset(tidyDf,type == 2),
                   alpha=0.2, aes(y=..density..), binwidth = .2, 
                   color=clrs[2], fill=clrs[2]) +
    geom_rug(alpha=0.3, show.legend=FALSE) +
    scale_fill_manual(values=clrs) + 
    scale_color_manual(values=clrs) + 
    theme(axis.title=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.line.y=element_blank(),
          legend.position="none",
          text = element_text(size=35, color="black"),
          plot.title=element_text(face="italic")) + coord_fixed(ratio=15) + 
    xlim(c(-0.1, 10)) +
    ylim(c(0,.63))+
    ggtitle(title) + 
    stat_function(fun="dnorm", colour=clrs[1],
                  args=list(mean=mean(exprNml[isof,]),
                            sd=sd(exprNml[isof,])), linetype="dashed", size=1.5) +
    geom_vline(size=2, alpha=0.8, color=clrs[1], xintercept = twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
    print(p1)
}



plotGene_bothHist <- function(exprNml, exprTum, isof, title, clrs){

  tidyDf <- as.data.frame(cbind(as.numeric(c(exprTum[isof,], exprNml[isof,])),
                                as.factor(c(rep("tumor",ncol(exprTum)), rep("normal",ncol(exprNml))))),
                          stringsAsFactors=FALSE)
  colnames(tidyDf) <- c("expr", "type")
  print(mean(exprNml[isof,]))
  print(twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
  expr <- type <- ..density.. <- NULL # Setting the variables to NULL first
  p1 <- ggplot(tidyDf, aes(x=expr, color=as.factor(type),
                           fill=as.factor(type),
                           group=as.factor(type))) +
    theme_classic() +
    geom_histogram(data=subset(tidyDf,type == 1),
                   alpha=0.2, aes(y=..density..), binwidth = .2, 
                   color=clrs[1], fill=clrs[1]) +
    geom_histogram(data=subset(tidyDf,type == 2),
                   alpha=0.2, aes(y=..density..), binwidth = .2, 
                   color=clrs[2], fill=clrs[2]) +
    geom_rug(alpha=0.3, show.legend=FALSE) +
    scale_fill_manual(values=clrs) + 
    scale_color_manual(values=clrs) + 
    theme(axis.title=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.line.y=element_blank(),
          legend.position="none",
          text = element_text(size=35, color="black"),
          plot.title=element_text(face="italic")) + coord_fixed(ratio=15) + 
    xlim(c(-0.1, 10)) +
    ggtitle(title) + 
    geom_vline(size=2, alpha=0.8, color=clrs[1], xintercept = twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
  print(p1)
}


plotGene_oneHist <- function(exprNml, exprTum, isof, title, clrs, vert_line = TRUE){
  
  tidyDf <- as.data.frame(cbind(as.numeric(c(exprTum[isof,], exprNml[isof,])),
                                as.factor(c(rep("tumor",ncol(exprTum)), rep("normal",ncol(exprNml))))),
                          stringsAsFactors=FALSE)
  colnames(tidyDf) <- c("expr", "type")
  print(mean(exprNml[isof,]))
  print(twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
  expr <- type <- ..density.. <- NULL # Setting the variables to NULL first
  p1 <- ggplot(tidyDf, aes(x=expr, color=as.factor(type),
                           fill=as.factor(type),
                           group=as.factor(type))) +
    theme_classic() +
    geom_histogram(data=subset(tidyDf,type == 1),
                   alpha=0.2, aes(y=..density..), binwidth = .2, 
                   color=clrs[1], fill=clrs[1]) +
     scale_fill_manual(values=clrs) + 
    scale_color_manual(values=clrs) + 
    theme(axis.title=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.line.y=element_blank(),
          legend.position="none",
          text = element_text(size=35, color="black"),
          plot.title=element_text(face="italic")) + coord_fixed(ratio=15) + 
    xlim(c(-0.1, 10)) +
    ggtitle(title) #+ 
  if(vert_line == TRUE){
    p1 <- p1 + geom_vline(size=2, alpha=0.8, color=clrs[1], xintercept = twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
  } 
  print(p1)
}


plotGene_oneHistNmlCurve <- function(exprNml, exprTum, isof, title, clrs, vert_line = TRUE, histogram = TRUE){
  
  tidyDf <- as.data.frame(cbind(as.numeric(c(exprTum[isof,], exprNml[isof,])),
                                as.factor(c(rep("tumor",ncol(exprTum)), rep("normal",ncol(exprNml))))),
                          stringsAsFactors=FALSE)
  colnames(tidyDf) <- c("expr", "type")
  print(mean(exprNml[isof,]))
  print(twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
  expr <- type <- ..density.. <- NULL # Setting the variables to NULL first
  p1 <- ggplot(tidyDf, aes(x=expr, color=as.factor(type),
                           fill=as.factor(type),
                           group=as.factor(type))) +
    theme_classic() +
    scale_fill_manual(values=clrs) + 
    scale_color_manual(values=clrs) + 
    theme(axis.title=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.line.y=element_blank(),
          legend.position="none",
          text = element_text(size=35, color="black"),
          plot.title=element_text(face="italic")) + coord_fixed(ratio=15) + 
    xlim(c(-0.1, 10)) +
    ggtitle(title) + 
    stat_function(fun="dnorm", colour=clrs[1],
                  args=list(mean=mean(exprNml[isof,]),
                            sd=sd(exprNml[isof,])), linetype="dashed", size=1.5)
    
  if(histogram == TRUE){
    p1 <- p1 + geom_histogram(data=subset(tidyDf,type == 1),
                              alpha=0.2, aes(y=..density..), binwidth = .2, 
                              color=clrs[1], fill=clrs[1])
  }
  if(vert_line == TRUE){
    p1 <- p1 + geom_vline(size=2, alpha=0.8, color=clrs[1], xintercept = twoSdBelowMean(mn=mean(exprNml[isof,]),sd = sd(exprNml[isof,])))
  }

  print(p1)
}




pvalFxn <- function(survDifObj){
  p.val <- 1 - pchisq(survDifObj$chisq, length(survDifObj$n) - 1)
  return(p.val)
}



sig2star = function(s, breaks=c(-Inf, 0.0001,0.001,0.01,1), labels=c("***","**","*","")) {
  r <- s
  r[] <- as.character(cut(s, breaks=breaks, labels)) 
  r[is.na(r)] <- ""
  return(r)
}



#' Cluster and Reorder Rows and Columns
#'
#' Clusters the rows and columns of a numeric matrix using hclust
#'
#' @param matr matrix to be clustered
#' @param dimen Indicates the dimension that the clustering is performed in. One of either "row", "column", or "both". 
#'
#' @return returns a matrix of clustered values by either row, column, or both, 

clusterAndReorder <- function(matr, dimen="both"){
  #dimen can be one of row, column or both
  if(dimen == "row"){
    hcl <- hclust(dist(matr))
    matr.new <- matr[hcl$order,]
    return(matr.new)
  } else if(dimen == "column"){
    hcl <- hclust(dist(t(matr)))
    matr.new <- matr[,hcl$order]
    return(matr.new)
  } else if(dimen == "both"){
    hcl.r <- hclust(dist(matr))
    hcl.c <- hclust(dist(t(matr)))
    matr.new <- matr[hcl.r$order, hcl.c$order]
    return(matr.new)
  }
}



set2 <- RColorBrewer::brewer.pal(n = 8, "Set2")
bw <- RColorBrewer::brewer.pal(n = 9, "Greys")

ggBar <- list(
  theme_classic(),
  theme(legend.position="top",
        axis.ticks = element_blank(),
        axis.text = element_text(size=18, color="black"),
        axis.text.x = element_text(angle = 90, hjust=1,vjust=0.5),
        axis.title = element_text(size=22)),
  scale_y_continuous(expand = c(0,0)))

ggHist <- list(
  theme_classic(),
  theme(legend.position="top",
        axis.text = element_text(size=24)),
  scale_y_continuous(expand = c(0,0)))


ggSeqPlate = list(
  geom_bar(stat="identity"),
  facet_grid(.~ensembl_gene_id),
  theme_classic(), 
  theme(axis.text=element_text(size=10),
        axis.text.x = element_text(angle=90, hjust = 1, vjust=0.5),
        axis.text.y = element_text(size=16)),
  scale_fill_manual(values=set2[2:1]))


clean_dataset <- function(e){
  
  #returns adj normal and tumor dataframes
  print(dim(e))
  
  #remove ffpe samples
  e <- e[,!e$is_ffpe]
  print("Dimensions after FFPE removed:")
  print(dim(e)) 
  
  #add technical variables
  colData(e)$hosp <- substr(rownames(colData(e)), 6,7)
  colData(e)$seqPlate <- substr(rownames(colData(e)), 22,25)
  
  ###remove all metaData columns that are NA
  colData(e) <- colData(e)[,colSums(is.na(colData(e)))<nrow(colData(e))]
  
  ##TPM normalization, then log2(tpm+1) transform
  eu <- assay(e)
  eu <- scale(eu, center=FALSE, scale=colSums(eu)) * 1000000
  eu <- log(eu+1, 2)
  assay(e) <- eu
  
  #separate tumor and adjacent normal samples
  et <- e[,substr(colnames(e),14,15) == "01"] 
  en <- e[,substr(colnames(e),14,15) == "11"]
  print("Dimensions of tumor and adj nml dataframes:")
  print(dim(et)); print(dim(en))
  
  #remove duplicates
  et <- et[,!duplicated(substr(colnames(et),1,12))] 
  en <- en[,!duplicated(substr(colnames(en),1,12))] 
  print("Dimensions of tumor and adj nml dataframes (duplicates removed):")
  print(dim(et)); print(dim(en))

  #remove patients with NA values
  et <- et[,!apply(assay(et), 2, anyNA)] 
  en <- en[,!apply(assay(en), 2, anyNA)] 
  print("Dimensions of tumor and adj nml dataframes (duplicates removed + pts with NA values removed):")
  print(dim(et)); print(dim(en))
  
  #Filter out genes where <20% of patients have a non zero expression value
  et <- et[rowSums(assay(et)==0)<=ncol(et)*.80,] #select genes w/ less than 80% zeros
  en <- en[rowSums(assay(en)==0)<=ncol(en)*.80,] #select genes w/ less than 80% zeros
  print("Dimensions of tumor and adj nml dataframes (duplicates removed + pts with NA values removed + 20% filter):")
  print(dim(et)); print(dim(en))
  
  #get the genes in common between the two
  genes_in_common <- intersect(rownames(et), rownames(en))
  et <- et[genes_in_common,]
  en <- en[genes_in_common,]
  print("Dimensions of tumor and adj nml dataframes (duplicates removed + pts with NA values removed + 20% filter + intersection):")
  print(dim(et)); print(dim(en))
  
  return(list(et=et, en=en))
}


centroid_plot_elem <- function(){
  list(ggplot2::theme_classic(), 
         geom_point(size=3),
         geom_text(aes(label=Type)))
}

returnDeMirnaLimma <- function(expr, design, colOfInt, pval=1, logFC=0, retAll=F){
  fit_t <- lmFit(object = expr, design = design)
  ebfit_t <- eBayes(fit_t)
  if(retAll == T){
    topTableRes_t <- topTable(ebfit_t, coef = colOfInt, number = Inf) 
    return(topTableRes_t)
  }
  
  topTableRes_t <- topTable(ebfit_t, coef = colOfInt, number = nrow(assay(expr)), adjust.method = "BH", p.value = pval, lfc = logFC) 
  #topTableRes_t2 <-topTableRes_t[{topTableRes_t$logFC > logFC} | {topTableRes_t$logFC < logFC},] 
  return(topTableRes_t)
}

assay_to_df <- function(sum_expt, tum_type, gene){
  data.frame(assay(sum_expt[gene,])) %>% 
    t %>% 
    as_tibble() %>%
    mutate(tum_type=tum_type)
}   


returnClassif <- function(tum_matr, tum_res, gene){
  res <- {assay(tum_matr[gene,]) > {tum_res %>% filter(geneNm == gene) %>% pull(lowerBound)}} %>%
    ifelse(1,0) %>% 
    as.data.frame() %>% 
    t() %>% 
    as.data.frame() %>% rename(!!paste0("classif_", gene) := gene)
  return(res)
}

calc_days_last_fu <- function(days_to_death, days_to_last_follow_up){
  days_last_fu <- ifelse(is.na(days_to_death), days_to_last_follow_up, days_to_death)
  days_last_fu <- ifelse(days_last_fu < 1, 1, days_last_fu)
  days_last_fu <- ifelse(days_last_fu > 365.25 * 5, 365.25 * 5, days_last_fu)
  days_last_fu <- as.data.frame(days_last_fu)
  return(days_last_fu)
}

calc_vital_stat <- function(vital_stat, days_last_fu ){
  vital_stat <- ifelse(vital_stat == "alive", 0, 1)
  vital_stat <- ifelse(days_last_fu >= 365.25 * 5, 0, vital_stat)
  vital_stat <- as.data.frame(vital_stat)
  return(vital_stat)
}

```

